---
title: "GGB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GGB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DDMethods)
library(data.table)
```

## General Growth Balance (GGB)

This vignette describes the methods for the General Growth Balance Death Distribution Method (DDM) for estimating completeness of death registration.
These methods are not original to this package, but are summarized here.
For original sources, see the references section of this vignette.

## Input data

Input data required for GGB is:

- Age-specific population counts from two censuses (recommended < 20 years apart)
- Age-specific death counts from death registration in between the two censuses

Age-specific death counts can come from one or many years of death registration, but should be converted to average annual counts.

This package includes an example dataset, which matches the [example used but the IUSSP](http://demographicestimation.iussp.org/content/generalized-growth-balance-method).

The example dataset is for South African males, between 2001 and 2007:

```{r zaf_data}
head(zaf_2001_2007)
```

As-is, these data are for total deaths in the interval, so we need to divide deaths by the time between censuses in years.
Note that in the `ggb` function, this happens internally, via the `input_deaths_annual` argument.

```{r zaf_annual}
# convert to average annual values
dt <- copy(zaf_2001_2007)
dt[, t := as.numeric(difftime(date2, date1, units = "days")) / 365]
dt[, deaths := deaths / t]
dt[, migrants := migrants / t]
```

## The Demographic Balancing Equation

At the core of the GGB method is the demographics balancing equation,
which says that in a stable population (constant birth rate and age-specific mortality rate over time),
population growth rate ($r$) is equal to birth rate ($b$) minus death rate ($d$) plus migration rate ($m$).
In a closed population, this simplifies to: $r = b - d$.

This base form of the demographic balancing equation looks at all ages and uses birth rate as the entry rate into the population.
But, we can think of an analogous format for any age $a$,
where "births" become birthdays at age $a$ and the population of interest is the population aged $a+$.
Then, $r(a+) = b(a+) - d(a+)$.

From two censuses, we can approximate $r(a+)$ and $b(a+)$ for a collection of ages $a$.
True death rate, $d(a+)$, itself is a function of observed death count, observed population counts, death registration completeness, and census coverage.
We leverage these relationships, along with orthogonal regression to summarize across ages, to estimate completeness.
The following sections will describe each component in more detail.

## A bit on notation

As described, we notate $r(a+)$, $b(a+)$, $d(a+)$, and $m(a+)$ to be the growth, birth (or entry), death, and migration rates in the population aged $a$ and over.

Additionally:

- $n$ : length of age interval in data
- $N1$ : population at census 1
- $N2$ : population at census 2
- $N$ : mean population between census 1 and census 2
- $D$ : average annual deaths recorded between censuses

And we use traditional demographic notation where ${_nZ_x}$ stands for some demographic measure, $Z$, for the age interval $x$ to $x+n$.

## Entry rate, b(a+)

We can estimate the entry rate into the population age $a+$ such that:
$$b(a+) = \frac{\text{birthdays aged } a}{N(a+)}$$
Where average annual birthdays aged $a$ is approximated as the geometric mean of ${_nN1_{a-n}}$ and ${_nN2_a}$ divided by the length of the age interval ($n$):
$$\text{birthdays aged } a = \frac{1}{n}\sqrt{{_nN1_{a-n}} \cdot {_nN2_a}}$$
And $N(a+)$ is approximated as the geometric mean of $N1(a+)$ and $N2(a+)$:
$$N(a+) = \sqrt{N1(a+) \cdot N2(a+)}$$
Note that if we assume census coverage is the same for all age groups, coverage does not impact this entry rate term.
This is because a coverage scalar applied to $N1$ and $N2$ would cancel in the numerator and denominator when we divide birthdays by $N(a+)$.

On our example dataset, we get:

```{r entry, echo=F}
results <- ggb(
  dt,
  age_trim_lower = 5,
  age_trim_upper = 85,
  id_cols = c("location", "sex", "age_start"),
  migration = FALSE,
  input_deaths_annual = TRUE,
  input_migrants_annual = TRUE
)
dt <- results[["dt"]]
base_cols <- c("age_start", "pop1", "pop2", "deaths")
print(dt[, .SD, .SDcols = c(base_cols, "bdays_age_a", "pop_age_aplus", "entry_rate")])
```


## Growth rate, r(a+)

The growth rate can be approximated as:

$$r(a+) = \frac{1}{t} \text{ln} \Bigg(\frac{N2(a+)}{N1(a+)}\Bigg)$$
Alternatively, it can be approximated as:

$$r(a+) = \frac{1}{t} \cdot \frac{N2(a+) - N1(a+)} {N(a+)}$$
with $N(a+)$ computed as it is described above.

The first method for estimating growth rate comes from Hill, You, and Choi (2009).
The second is used by IUSSP and [Tim Riffe's DDM package](https://github.com/timriffe/AdultCoverage/tree/master/AdultCoverage/R/DDM).

Now, consider the case where census 1 and census 2 have coverage $k1$ and $k2$ such that true populations are $\frac{N1(a+)}{k1}$ and $\frac{N2(a+)}{k2}$.
Coverage is assumed to be the same for all age groups.
Let $r(a+)$ be growth rate as estimated in the first equation above, and
let $r(a+)'$ be growth rate that incorporates $k1$ and $k2$.
Using the first growth equation, we get:
$$r(a+)' = \frac{1}{t} \text{ln} \Bigg(\frac{N2(a+) / k2}{N1(a+) / k1}\Bigg)$$
when we apply coverage scalars to the observed $N$ values.

With some algebra, we can get:

$$ \implies r(a+)' = \frac{1}{t} \Bigg( \text{ln} \Bigg(\frac{N2(a+)}{N1(a+)} \Bigg) + \text{ln}\Big(\frac{k1}{k2} \Big)\Bigg) $$
$$ \implies r(a+)' = r(a+) + \delta \text{ where } \delta = \frac{1}{t}\text{ln}\big(\frac{k1}{k2}\big)$$
and we can think of $\delta$ here as the error in growth rate when estimated using $N2(a+)$ and $N1(a+)$.
We will use this $\delta$ later in the demographic balancing equation.

In our example dataset, we find growth rate ($r(a+)$) to be:

```{r growth, echo=F}
print(dt[, .SD, .SDcols = c(base_cols, "growth_rate")])
```

## Death rate, d(a+)

Given that death rate is by definition deaths over person-years, we can get an incomplete death rate as follows:

$$d(a+) = \frac{D(a+)}{N(a+)}$$

where the numerator is deaths from the data and the denominator is population as estimated above.

When death registration is incomplete, true deaths are $\frac{D(a+)}{c}$ where $c$ is death registration completeness.
Additionally, the denominator $N(a+)$ is estimated from the two censuses and may be biased if one census has higher coverage.

To estimate this bias from censuses, we'll go back to how we estimated $N(a+)$,
apply $k1$ and $k2$ scalars to census 1 and census 2,
and see how that modifies our population estimate.
Let $N(a+)'$ be a version of population that accounts for $k1$ and $k2$ and $N(a+)$ remain population as estimated above.
So:

$$N(a+)' = \sqrt{\frac{N1(a+) \cdot N2(a+)}{k1 \cdot k2}}$$

$$ \implies N(a+)' = \Big(\sqrt{N1(a+) \cdot N2(a+)} \Big) \Big(\sqrt{\frac{1}{k1 \cdot k2}} \Big)$$

$$ \implies N(a+)' = N(a+) \cdot \frac{1}{\sqrt{k1 \cdot k2}}$$

Next, go back to the formulation of $d(a+)$.
Substitute $\frac{D(a+)}{c}$ for the numerator and substitute $N(a+)'$ for $N(a+)$ in the denominator.
Doing this, we get:

$$d(a+) = \frac{\frac{D(a+)}{c}}{\frac{N(a+)}{(k1 \cdot k2) ^ \frac{1}{2}}}$$

$$ \implies d(a+) = \frac{(k1 \cdot k2) ^ \frac{1}{2}}{c} \cdot \frac{D(a+)}{N(a+)}$$

## Regression to estimate completeness

Putting this together by using the balancing equation, and error components for growth rate and death rate as described, we can evaluate:

$$b(a+) - r(a+) = d(a+)$$
$$\implies b(a+) - \Big[r(a+) + \delta \Big] = \frac{(k1 \cdot k2) ^ \frac{1}{2}}{c} \cdot \frac{D(a+)}{N(a+)}$$

$$ \implies b(a+) - r(a+) = \text{ intercept } + \text{ slope } \cdot \frac{D(a+)}{N(a+)}$$

Where:

$$\text{intercept } = \delta = \frac{1}{t} \text{ln} \Big(\frac{k1}{k2} \Big)$$

$$\text{slope } = \frac{(k1 \cdot k2)^\frac{1}{2}}{c}$$
At this point, the intercept and slope are the only unknowns, so this is a system we know how to solve.
We use all values of $a$ within a chosen range ("age trim"), combined with orthogonal regression, to estimate slope and intercept.
Orthogonal regression for $y \sim x$ has an analytical solution as follows:

$$\text{slope } = \frac{\text{sd}(y)}{\text{sd}(x)}$$

$$\text{intercept } = \text{mean}(y) - \text{mean}(x) \cdot \text{ slope }$$
Then, solve for completeness using slope, intercept, and the relationships above, as follows:

1. Solve for relative census coverage:
$$\text{intercept } = \frac{1}{t} \text{ln} \Big(\frac{k1}{k2} \Big)$$
$$\implies {\frac{k1}{k2} = \text{exp}(\text{intercept} \cdot t)}$$

2. Use $\frac{k1}{k2}$ to select $k1$ and $k2$ assuming the more complete census is 100% complete
3. Then: $$c = \frac{(k1 \cdot k2)^\frac{1}{2}}{\text{slope}}$$

## Using this package for GGB

To use this package for GGB, use the `ggb` function:

```{r example}
results <- ggb(
  dt,
  age_trim_lower = 5,
  age_trim_upper = 85,
  id_cols = c("location", "sex", "age_start"),
  migration = FALSE,
  input_deaths_annual = TRUE
)
```

This will return a list of two objects.
The first has your input data with GGB component columns added:

```{r example_outputs1}
head(results[["dt"]])
```

The second has completeness by each set of ID columns, as well as summaries such as slope,
intercept, and relative census coverage:

```{r example_outputs2}
print(results[["completeness"]])
```

You can also plot the linear fit using the `plot_ggb` function:

```{r plot_results, fig.width=7,fig.height=5}
id_cols_subset = list(
  location = "South Africa",
  sex = "male",
  age_start = seq(5, 80, 5)
)
gg <- plot_ggb(results, id_cols_subset = id_cols_subset)
print(gg)
```

## References

 * Bennett NG, Horiuchi S. Mortality estimation from registered deaths in
   less developed countries. Demography. 1984 May;21(2):217-33.
 * Dorrington RE. 2013. "The Generalized Growth Balance Method".
   In Moultrie TA, RE Dorrington, AG Hill, K Hill, IM Timæus and B Zaba (eds).
   Tools for Demographic Estimation. Paris: International Union for the
   Scientific Study of Population.
   http://demographicestimation.iussp.org/content/generalized-growth-balance-method.
   Accessed May 27 2021.
 * Hill K, You D, Choi Y. Death distribution methods for estimating adult
   mortality: sensitivity analysis with simulated data errors. Demographic
   Research. 2009 Jul 1;21:235-54.
 * Murray CJ, Rajaratnam JK, Marcus J, Laakso T, Lopez AD. What can we
   conclude from death registration? Improved methods for evaluating
   completeness. PLoS Med. 2010 Apr 13;7(4):e1000262.
 * Tim Riffe, Everton Lima, and Bernardo Queiroz (2017). DDM: Death
   Registration Coverage Estimation. R package version 1.0.0.
   https://CRAN.R-project.org/package=DDM
