---
title: "GGBSEG"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GGBSEG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DDMethods)
library(data.table)
```

## General Growth Balance - Synthetic Extinct Generations Hybrid (GGBSEG)

This vignette describes the methods for the General Growth Balance - Synthetic Extinct Generations hybrid Death Distribution Method (DDM) for estimating completeness of death registration.
These methods are not original to this package, but are summarized here.
For original sources, see the references section of this vignette.

This vignette also builds off of the vignettes for the GGB and SEG methods.

## Input data

See the sections on input data in the GGB and SEG vignettes.

## The GGBSEG method

The GGBSEG method combines GGB and SEG as follows:

1. The first step of the GGBSEG hybrid method is to run the GGB method.

2. One output of GGB is the coverage of census 1 relative to census 2 ($\frac{k1}{k2}$).
Population counts from census 1 are divided by this relative coverage value to get adjusted census 1 population.

3. Finally, the adjusted census 1 population, along with the original input for deaths and census 2 population, are used as input to the SEG method.
The output of SEG is the final completeness estimate for GGBSEG.

Note that different age trims can be used for the GGB component and the SEG component.

## Using this package for GGBSEG

To use this package for GGBSEG, use the `ggbseg` function:

```{r example}
# setup data
data(zaf_2001_2007)
dt <- copy(zaf_2001_2007)
dt$cd_region <- "west"

# run ggbseg
results <- ggbseg(
  dt,
  age_trim_lower_ggb = 5,
  age_trim_upper_ggb = 75,
  age_trim_lower_seg = 55,
  age_trim_upper_seg = 90, 
  id_cols = c("location", "sex", "age_start"),
  migration = FALSE,
  input_deaths_annual = FALSE
)
```

This will return a list of two objects.
The first has your input data with intermediate columns added:

```{r example_outputs1}
head(results[["dt"]])
```

The second has completeness by each set of ID columns:

```{r example_outputs2}
print(results[["completeness"]])
```

## References

 * Bennett NG, Horiuchi S. Mortality estimation from registered deaths in
   less developed countries. Demography. 1984 May;21(2):217-33.
 * Dorrington RE. 2013. "The Generalized Growth Balance Method".
   In Moultrie TA, RE Dorrington, AG Hill, K Hill, IM Timæus and B Zaba (eds).
   Tools for Demographic Estimation. Paris: International Union for the
   Scientific Study of Population.
   http://demographicestimation.iussp.org/content/generalized-growth-balance-method.
   Accessed May 27 2021.
 * Hill K, You D, Choi Y. Death distribution methods for estimating adult
   mortality: sensitivity analysis with simulated data errors. Demographic
   Research. 2009 Jul 1;21:235-54.
 * Murray CJ, Rajaratnam JK, Marcus J, Laakso T, Lopez AD. What can we
   conclude from death registration? Improved methods for evaluating
   completeness. PLoS Med. 2010 Apr 13;7(4):e1000262.
 * Tim Riffe, Everton Lima, and Bernardo Queiroz (2017). DDM: Death
   Registration Coverage Estimation. R package version 1.0.0.
   https://CRAN.R-project.org/package=DDM
