---
title: "SEG"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DDMethods)
library(data.table)
```

## Synthetic Extinct Generations (SEG)

This vignette describes the methods for the Synthetic Extinct Generations Death Distribution Method (DDM) for estimating completeness of death registration.
These methods are not original to this package, but are summarized here.
For original sources, see the references section of this vignette.

## Input data

Input data required for SEG is:

- Age-specific population counts from two censuses (recommended < 20 years apart)
- Age-specific death counts from death registration in between the two censuses

Age-specific death counts can come from one or many years of death registration, but should be converted to average annual counts.

This package includes an example dataset, which matches the [example used but the IUSSP](http://demographicestimation.iussp.org/content/generalized-growth-balance-method).

The example dataset is for South African males, between 2001 and 2007:

```{r zaf_data}
head(zaf_2001_2007)
```

As-is, these data are for total deaths in the interval, so we need to divide deaths by the time between censuses in years.
Note that in the `seg` function, this happens internally, via the `input_deaths_annual` argument.

```{r zaf_annual}
# convert to average annual values
dt <- copy(zaf_2001_2007)
dt[, t := as.numeric(difftime(date2, date1, units = "days")) / 365]
dt[, deaths := deaths / t]
dt[, migrants := migrants / t]
```

## A bit on notation

Assume the following notation:

- $n$ : length of age interval in data
- $N1$ : population at census 1
- $N2$ : population at census 2
- $N$ : mean population between census 1 and census 2
- $D$ : average annual deaths recorded between censuses
- $_nr_x$ : age-specific growth rate between ages $x$ and $x+n$
- $r_x$ : cumulative growth rate up to age $x$
- $e_x$ : life expectancy at age $x$

Additionally, we use traditional demographic notation where ${_nZ_x}$ stands for some demographic measure, $Z$, for the age interval $x$ to $x+n$.

## Summary

Consider a hypothetical birth cohort, born in 1900, with a population of 100 at age 50 in the year 1950.
Assuming there is no migration and the vital registration is complete (and no one lived beyond age 120),
we could count up all of the deaths observed in this birth cohort between 1950 and present-day and it should add up to 100 deaths.
In other words, each of the 100 individuals in the population in 1950 died and was counted by the vital registration system.
If instead, 90 deaths were counted, we could compare the 90 recorded deaths to the population of 100 and estimate completeness to be 90%.
The SEG method leverages this relationship between population and mortality,
but uses period-specific mortality as opposed to cohort-specific mortality.

To summarize the basic relationship:
$$\text{completeness} \approx \frac{\text{pop estimated from deaths}}{\text{pop estimated from censuses}}$$
Note that the population of interest in the numerator and denominator can be:

1. $N_x$ : the population aged $x$, also equivalent to the average number of birthdays age $x$ in one year.
2. ${_nN_x}$ : the population in the age interval $x$ to $x+n$ where $n$ is commonly 5 to match 5-year age intervals present in abridged data.

Both numerator and denominator types can be used in the `DDMethods` package.

## Population from censuses

The population from censuses can be estimated as follows:

$N_a$: Average annual birthdays aged $a$ is approximated as the geometric mean of ${_nN1_{a-n}}$ and ${_nN2_a}$ divided by the length of the age interval ($n$):
$$\text{birthdays aged } a = \frac{1}{n}\sqrt{{_nN1_{a-n}} \cdot {_nN2_a}}$$
$_nN_a$: Population aged $a$ to $a+n$ is approximated as the geometric mean of $_nN_a$ from census 1 and $_nN_a$ from census 2:
$$_nN_a = \sqrt{_nN1_a \cdot _nN2_a}$$

## Population from death registration

The population from death registration is estimated using the synthetic cohort method.
In short, we first compute age-specific population growth between censuses,
then aggregate to get cumulative growth over age.
Cumulative growth rate is used to go from recorded deaths to $_nd_x$ to get the ratio of $_{30}d_{10}$ to $_{20}d_{40}$.
This ratio is used to match with Coale-Demeny model life tables, to approximate life expectancy in the open age interval.
From here, the synthetic cohort population is computed recursively starting at the open interval and working backwards,
using recorded deaths, life expectancy, and age-specific growth rate.
The following sections will describe each of these steps in more detail.

### Age-specific population growth

For the age group in interval $a$ to $a+n$,
age-specific population growth, $_nr_a$, can be approximated as:

$$_nr_a = \frac{1}{t} \cdot \text{ln} \Big( \frac{_nN2_a}{nN1_a} \Big)$$

### Cumulative growth over age

To get cumulative growth over age, compute:
$$r_a = n \cdot \sum_{x = 0}^{a-n}(_nr_x) + \big(\frac{n}{2} \cdot {_nr_a}\big)$$

according to Bennett and Horiuchi (1984).

### $_nd_x$ and ratio $_{30}d_{10}$ to $_{20}d_{40}$

Then, $_nd_x$ is equal to recorded deaths times the exponentiated cumulative growth rate,
again following Bennett and Horiuchi:

$$_nd_x = {_nD_x} \cdot \text{exp}(r_x)$$

From here it is straightforward to sum over the relevant age groups and divide to get $\frac{_{30}d_{10}}{_{20}d_{40}}$.

### Life expectancy in the open interval

Using the Coale Demeny model life tables,
we can construct a function that maps from $\frac{_{30}d_{10}}{_{20}d_{40}}$ to life expectancy for a given open age interval.

This plot shows the relationship between $\frac{_{30}d_{10}}{_{20}d_{40}}$ and life expectancy for the CD West male life tables:

```{r cdwest, echo=F, fig.width=7,fig.height=5}
library(ggplot2)
cd <- copy(coale_demeny_ex)
cd <- cd[sex == "male" & cd_region == "west"]

gg <- ggplot(data = cd[age_start >= 60],
             aes(x = ratio_30d10_20d40,
                 y = ex,
                 color = as.factor(age_start))) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(x = "Ratio 30d10 to 20d40",
       y = "Life expectancy (ex)",
       color = "Open age group",
       title = "Coale Demeny West, males") +
  scale_x_continuous(breaks = seq(0, 2, 0.2)) +
  scale_y_continuous(breaks = seq(0, 20, 5))
print(gg)
```

The `DDMethods` package uses linear interpolation of ex given discrete values from Coale Demeny life tables to get a continuous mapping from ratio to ex.

The `seg` function requires a `cd_region` column in your input data to select which Coale Demeny region should be used for this mapping.
The options are 'west', 'east', 'south', and 'north'.

```{r cd_region}
dt[, cd_region := "west"]
```
  

### Population

Next, start with the open age interval and find:

$$N(a)_{\text{open}} = {_nD_a} \Bigg[ \text{exp}(e_a \cdot {_nr_a}) - \frac{(e_a \cdot {_nr_a})^2 }{6} \Bigg]$$
   
Then, recursively compute population for previous age groups:

$$N(a) = N(a+n) \cdot \text{exp}(n \cdot {_nr_a}) + {_nD_a} \cdot \text{exp}(0.5 \cdot n \cdot {_nr_a})$$
To use the population numerator and denominator type $_nN_x$,
further calculate:

$$_nN_a = (N(a) + N(a+n)) (\frac{n}{2})$$

For more details on derivation of these equations, see the references below.

## Synthesis to estimate completeness

For each age, $a$, within the selected age trim interval, compute:

$$\text{completeness} \approx \frac{\text{pop estimated from deaths}}{\text{pop estimated from censuses}}$$
And then take the mean over the completeness estimates generated from this set of ages to get a final estimate of completeness.


## Using this package for SEG

To use this package for SEG, use the `seg` function:

```{r example}
results <- seg(
  dt,
  age_trim_lower = 55,
  age_trim_upper = 90,
  id_cols = c("location", "sex", "age_start"),
  migration = FALSE,
  input_deaths_annual = TRUE,
  method_numerator_denominator_type = "Nx"
)
```

This will return a list of two objects.
The first has your input data with SEG component columns added:

```{r example_outputs1}
head(results[["dt"]])
```

The second has completeness by each set of ID columns, as well as summaries such as slope,
intercept, and relative census coverage:

```{r example_outputs2}
print(results[["completeness"]])
```

To evaluate completeness from each component age value, $a$, we can do the following:

```{r example_each_a, fig.width=7,fig.height=5}
dt <- results[["dt"]]
dt[, completeness := pop_from_deaths / pop_from_censuses]
dt[, included_in_age_trim := between(age_start, 55, 90)]
gg <- ggplot(data = dt[!is.na(completeness)],
             aes(x = age_start, y = completeness,
                 color = included_in_age_trim)) +
  geom_point() +
  geom_hline(yintercept = results[["completeness"]]$completeness) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 85, 5)) +
  labs(x = "Age (a)", y = "Completeness", title = "ZAF males 2007-2010",
       color = "Included in age-trim")
print(gg)
```

Note that in some cases, like here, you can see how age-trim selection would strongly influence the summary of completeness.
In our simulation exercises, we found that [55, 90) performed the best.
If the completeness has an increasing or decreasing trend with $a$,
it may be a sign that there is either strong migration or large variation in census coverage.
The later is a situation where the GGB-SEG hybrid is useful.

Additionally, we need to be careful in the interpretation of these plots to note that this is not age-specific completeness.
In other words, it is not completeness for the interval $a$ to $a+n$.
Rather, it is overall completeness estimated using information from ages $a$ and above.


## References

 * Bennett NG, Horiuchi S. Mortality estimation from registered deaths in
   less developed countries. Demography. 1984 May;21(2):217-33.
 * Dorrington RE. 2013. "Synthetic extinct generations methods".
   In Moultrie TA, RE Dorrington, AG Hill, K Hill, IM Timæus and B Zaba (eds).
   Tools for Demographic Estimation. Paris: International Union for the
   Scientific Study of Population.
   http://demographicestimation.iussp.org/content/synthetic-extinct-generations-methods.
   Accessed May 27 2021.
 * Hill K, You D, Choi Y. Death distribution methods for estimating adult
   mortality: sensitivity analysis with simulated data errors. Demographic
   Research. 2009 Jul 1;21:235-54.
 * Murray CJ, Rajaratnam JK, Marcus J, Laakso T, Lopez AD. What can we
   conclude from death registration? Improved methods for evaluating
   completeness. PLoS Med. 2010 Apr 13;7(4):e1000262.
 * Tim Riffe, Everton Lima, and Bernardo Queiroz (2017). DDM: Death
   Registration Coverage Estimation. R package version 1.0.0.
   https://CRAN.R-project.org/package=DDM

